DRAFT: Factors Influencing Abundance & Virus Isolation in Mosquitoes

### 2024-02-23 products/manuscript/Manuscript.qmd
Part 2
This commit includes the work associated with part 2 of the project including the processing code located here: /R/processing-code/MADAproject_pt2_ruiz.qmd

---
title: "Class Exercise: Factors Influencing Abundance & Virus Isolation in Mosquitoes"
subtitle: "Which environmental, ecological, and temporal factors impact mosquito populations  "
author: Andrew Ruiz
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
editor: 
  markdown: 
    wrap: 72
---

@@

Since 2000, there have been 45 human cases recorded in Massachusetts,
resulting in 22 deaths. There are no human vaccines for EEE, and
treatment consists of palliative care only. Prevention measures include
personal behaviors to avoid mosquito bites and decreasing the mosquito
population through pesticide use and environmental modifications like
removing standing water. The Bristol County Mosquito Control Project
(BCMCP) coordinates mosquito surveillance and testing in the county from
June to October. BCMCP has used the same sentinel collection sites and
trapping methods for over 20 years. Once trapped, mosquitoes are sorted
by species, and vector species are submitted to the Massachusetts
Department of Health’s (MDPH) State Lab for PCR virus testing. Results
are available within 24 hours of submission. When rates of EEE mosquito
infections are above a defined threshold, MDPH notifies the local boards
of health and recommends preventive measures. Recommended measures
include outreach and education to increase personal protective practices
(Heymann, 2015). Additionally, MDPH may recommend that cities and towns
in the affected areas cancel evening outdoor events and discourage
outdoor activities when mosquito vectors are most active (dawn and dusk)
(Massachusetts Department of Public Health, n.d.). When infection rates
are high enough to indicate imminent human transmission, the
Commonwealth of Massachusetts will recommend and fund aerial pesticide
applications over the affected areas.

@@

| Data                         | Description                                                                                                                                           | Source                                |
|------------------|-------------------------------|-----------------------|
| Mosquito Surveillance Data   | Records of mosquito species counts, locations, and dates of collection. Data on the presence of EEE in mosquito populations is particularly valuable. | XXX County Mosquito Control           |
| Weather Data                 | Temperature, humidity, rainfall, and wind speed influence mosquito activity, population dynamics, and virus transmission.                             | NOAA                                  |
| Bird Population Data         | Since birds are a natural reservoir for EEE, information on bird populations and migration patterns could be relevant.                                | Crowd-sourced bird data               |
| Human Case Reports           | Data on confirmed human cases of EEE, including location, date of onset, and clinical outcomes.                                                       | MA Dept of Public Health publications |
| Veterinary Surveillance Data | Since horses and other mammals can also be affected by EEE, veterinary records could provide early warning signs of virus activity in an area.        | MA Dept of Public Health              |

### Currently, I have collected 2 datasets that are described below:

#### The first dataset is the mosquito collection dataset

```{r}
dp1 = here("data", "raw-data", "MOSQ_MADA.xlsx")

mos_data = read_excel(dp1)

head(mos_data)

#### The second dataset is the weather data spanning the same time period as the mosquito collection data

```{r}
dp2 = here("data", "raw-data", "weather_airport.csv")
weather = read.csv(dp2, fileEncoding = "UTF-8")
head(weather)
```

@@

To cite other work (important everywhere, but likely happens first in
introduction), make sure your references are in the bibtex file
specified in the YAML header above (here
`dataanalysis_template_references.bib`) and have the right bibtex key.
Then you can include like this:

Examples of reproducible research projects can for instance be found in
[@mckay2020; @mckay2020a].

@@

*This section will be updated when all data sources are confirmed*

@@

*This step will be completed once all data sources are confirmed*

@@

*These will be generated in later phases of the exercise*

## Basic statistical analysis

## Full analysis

# Discussion

## Strengths and Limitations

## Conclusions

*Include citations in your Rmd file using bibtex, the list of references
will automatically be placed at the end*

This paper [@leek2015] discusses types of analyses.

These papers [@mckay2020; @mckay2020a] are good examples of papers
published using a fully reproducible setup similar to the one shown in
this template.

Note that this cited reference will show up at the end of the document,
the reference formatting is determined by the CSL file specified in the
YAML header. Many more style files for almost any journal [are
available](https://www.zotero.org/styles). You also specify the location
of your bibtex reference file in the YAML. You can call your reference
file anything you like, I just used the generic word `references.bib`
but giving it a more descriptive name is probably better.

@@

>>>>

---
title: "DRAFT: Factors Influencing Abundance & Virus Isolation in Mosquitoes"
subtitle: "Which environmental, ecological, and temporal factors impact mosquito populations  "
author: Andrew Ruiz
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
editor: 
  markdown: 
    wrap: 72
---

@@

Since 2000, there have been 45 human cases recorded in Massachusetts,
resulting in 22 deaths. There are no human vaccines for EEE, and
treatment consists of palliative care only. Prevention measures include
personal behaviors to avoid mosquito bites and decreasing the mosquito
population through pesticide use and environmental modifications like
removing standing water. The Bristol County Mosquito Control Project
(BCMCP) coordinates mosquito surveillance and testing in the county from
June to October. BCMCP has used the same sentinel collection sites and
trapping methods for over 20 years. Once trapped, mosquitoes are sorted
by species, and vector species are submitted to the Massachusetts
Department of Health’s (MDPH) State Lab for PCR virus testing. Results
are available within 24 hours of submission. When rates of EEE mosquito
infections are above a defined threshold, MDPH notifies the local boards
of health and recommends preventive measures. Recommended measures
include outreach and education to increase personal protective practices
. Additionally, MDPH may recommend that cities and towns in the affected
areas cancel evening outdoor events and discourage outdoor activities
when mosquito vectors are most active (dawn and dusk). When infection
rates are high enough to indicate imminent human transmission, the
Commonwealth of Massachusetts will recommend and fund aerial pesticide
applications over the affected areas.

@@

| Data                         | Description                                                                                                                                           | Source                                |
|------------------|-------------------------------|-----------------------|
| Mosquito Surveillance Data   | Records of mosquito species counts, locations, and dates of collection. Data on the presence of EEE in mosquito populations is particularly valuable. | XXX County Mosquito Control           |
| Weather Data                 | Temperature, humidity, rainfall, and wind speed influence mosquito activity, population dynamics, and virus transmission.                             | NOAA                                  |
| Bird Population Data         | Since birds are a natural reservoir for EEE, information on bird populations and migration patterns could be relevant.                                | eBird                                 |
| Human Case Reports           | Data on confirmed human cases of EEE, including location, date of onset, and clinical outcomes.                                                       | MA Dept of Public Health publications |
| Veterinary Surveillance Data | Since horses and other mammals can also be affected by EEE, veterinary records could provide early warning signs of virus activity in an area.        | MA Dept of Public Health              |

@@

@@

*This section will be updated before analysis begins*

@@

Datasets will be loaded into R in a Quarto document to document and
describe cleaning methods. There are two datasets that will require more
extensive cleaning: the mosquito trap data and the crowd-sourced bird
data. The process is outlined in detail in the MADAproject_pt2.qmd file.
Less significant cleaning and data management steps not included in this
manuscript are also outlined in that file.

The mosquito dataset spans 2007 to 2023. However, the virus and case
data only covers the years 2014 to 2020. Because mosquito populations
can be influences by processes of the previous year. We will set the
analytical horizon will span 2013 to 2021.

Additionally, the Massachusetts Department of Public Health requires
that all mosquito control district submit their colelction data in a
specific format. This format divides each trap event into separate rows
based on mosquito species. However, this format is not useful for
certain calculations where a zero count for a species is not explicitly
recorded. For example, even when sorted by species to calculate the
average count per trap, the average could be inflated since the records
will not include trap events with a zero count of that species. In order
to correct this, the data was pivoted to create a column for every
unique "species code" then the "pool size" for each species code is
summed for each trap event. A trap event is when "town", "date of
collection", "trap type" are all the same.. For species codes not
included in the orginal table for a trap event, the field will be
assigned a zero value. The resulting table will have one row for every
trap event.

The bird observation dataset was obtained from Cornell University's
eBird program. The original dataset included all reported bird
observations for Southeast Massachusetts and was over 600 MB of data.
While all birds are potential amplifying hosts for EEE and West Nile
Virus (WNV), birds from the Genus *Corvus* are of particular concern.
The bird dataset processed to include records of *Corvus spp* sightings
after 2014. The final dataset included 8 of the 50 variables, including
location data.

@@

The first phase of the exploration is complete. The next phase will
include statistical tests to determine if there are possible
associations between the environmental variables and the mosquito
abundance as measured by trap counts or EEE virus activity.

One factor to consider is if early virus isolations are associated with
cases in humans and other mammals. The table below shows the date of the
first EEE virus isolation in mosquitoes and the number of human of other
animal cases by year.

![](images/eee_isolation_cases-01.png){fig-align="center" width="320"}

Unfortunately, the time horizon is limited by data availability.
Additionally, EEE human infections are rare, so traditional statistical
methods may be inadequate to suggest any relationships. What is clear is
that 2019 was an exceptional year for EEE.

*Culiseta melanura* (MEL) plays an important role in EEE virus
amplification in the enzootic cycle. Understanding the population
dynamics might provide more insight into what is driving the 2019 spike.

The table below shows the proportion of MEL among all mosquito vector
species trapped in Bristol County, MA.

![](images/mel_proportion.png){fig-align="center"}

From this we see a large proportion of MEL in the traps for 2019. In
fact, they comprise more than half of all vector species for that year.
Another interesting aspect is the high proportion in 2018. MEL
overwinters as larvae and as long as there is sufficient standing water
to last the winter, MEL hatched in the previous fall would survive to
emerge the next year. This suggests it might be important to look at
conditions in the previous year or season for a better understanding.

Since water plays a significant role in the mosquito life cycle, let's
look at the annual precipitation.

![](images/annual_precip-02.png){fig-align="center"}

The graph indicates that the annual precipitation total for 2018 is
higher than the other years. This suggest that there was enough water to
support large MEL larval populations though out the winter.

## Basic statistical analysis
*These will be generated in later phases of the exercise*

## Fullanalysis

*These will be generated in later phases of the exercise* 
# Discussion *this section will be completed when analysis is done*

## Strengths and Limitations 
*this section will be updated after analysis is complete*

While the reliability of the mosquito trap data is high, the trap data
cannot be directly associated with the virus isolation data. The
information this scale is not released by the State. Spatial scale is
also an issue with the human and veterinary cases data. Human case data
is released only on a county scale to ensure patient privacy. Some
veterinary case data is available on the town-level, but most remains
only on the county-level.

These limitations are barriers to accurate spatial analysis.

## Conclusions *this section will be completed when analysis is done*
{{< pagebreak >}}

@@

References will be added in part 3 of the project
